Modelling last glacial cycle ice dynamics in the Alps

### 2017-09-19 articles/cycle/alps-cycle.tex
Rephrase introduction and fill logical gaps.
The five bullet points (extent, flow, timing, thickness, dynamics) should form
the red line of the paper. They correspond to subsections in the discussion,
and bullet points in the conclusions.

    Glaciers move by a combination of meltwater-induced sliding at the base
    \citep[\S532]{Saussure.1779}, and viscous deformation within the ice body
    \citep{Forbes.1846b}. As glaciers flow and slide across their bed, they
    transport rock debris and erode the landscape, thereby leaving
    geomorphological traces of their former presence.

    For nearly 300~years, montane people and early explorers of the European
    Alps learned to read the geomorphological imprint left by glaciers in the
    landscape, and to understand that glaciers had once been more extensive
    than today \citep[e.g.,][p.~21]{Windham.Martel.1744}. In the mid-nineteenth
    century, more systematic studies of glacial features showed that alpine
    glaciers extended well below their contemporary margins \citep{Venetz.1821}
    and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
    idea that, under colder temperatures, expansive ice sheets had once covered
    much of Europe and North America \citep{Agassiz.1840}.

    However, this glacial theory did not gain general acceptance until the
    discovery and exploration of the two present-day ice sheets on Earth, the
    Greenland and Antarctic ice sheets, provided a modern analogue for the
    proposed European and North American ice sheets. Although it was long
    unclear whether there had been a single or multiple glaciations, this
    controversy ended with the large-scale mapping of two distinct moraine
    systems in North America \citep{Chamberlin.1894}. In the European Alps, the
    systematic classification of the glaciofluvial terraces on the northern
    foreland indicated that landscape, that there had been at least four major
    glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
    of glaciofluvial stratigraphy in the foreland indicate up to eight
    glaciations \citep{Ivy-Ochs.etal.2008, Preusser.etal.2011}.

    More recently, palaeoclimate records extracted from deep sea sediments and
    ice cores have provided a much more detailed picture of the Earth
    environmental history \citep[e.g.,][]{Emiliani.1955,
    Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
    indicating neither one, nor four, but rather tens of glacial cycles. During
    the last 800000 years (800\,ka), glacial and interglacial periods have
    succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
    Augustin.etal.2004}. However, much of the traces left by glaciers on the
    landscape date from the most recent of these cycles, the \emph{last glacial
    cycle}.

    Nevertheless, the landform record is sparse in time and, most often,
    spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
    on the lanscape, and much of this evidence has been overprinted by
    subsequent glacier re-advances and other geomorphological processes
    \citep{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. The landscape
    imprint of previous glaciations has been largely overprinted by the more
    recent ones; therefore most of the glacial features currently left on the
    foreland present a record of the last major glaciation of the Alps, dating
    from the Last Glacial Maximum (LGM).

    Because this global signal is largely governed by the North American and
    Eurasian ice sheet complexes, it is unclear wether glacier advances and
    retreats in the Alps were in pac(71--57\,ka)e with global sea-level
    fluctuations. The timing of the LGM in the Alps (Ivy-Ochs et al., 2008;
    Monegato et al., 2007) is in good agreement with the maximum expansion of
    continental ice sheets recorded by the Marine Oxygen Isotope Stage (MIS) 2
    between 29 000 and 14 000 years before present (BP) (Lisiecki and Raymo,
    2005). There are nonetheless regional variations between different piedmont
    lobes, that are of the same order of magnitude with the dating
    uncertainties. Because recent glaciation overprinted older ones, and
    because dating uncertainties typically increase with age, dated
    reconstructions are strongly biased towards earlier glaciations
    \citep{Heyman.etal.2011}, and thus much of the glacial records dates to the
    Last Glacial Maximum (LGM). The Alps have been studied in more detail than
    any other glacial region. There is evidence for glaciation during MIS 4
    (71--57\,ka) but potential foreland glaciations during the rest of the last
    glacial cycle are still discussed.

    Here, we use the Parallel Ice Sheet Model
    \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
    approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
    model alpine glacier dynamics through the last glacial cycle (120--0\,ka),
    a period for which palaeo-temperature proxies are available, albeit non
    regional. We test the model sensitivity to multiple palaeo-climate forcing
    (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
    at high resolution for the optimal forcing (Sect.~\ref{sec:results}).

    \idea{Matthias: I think it is very important to emphasize the knowledge gap
          that you fill in with your work in the Intro. (1) it is incompletely
          known how many advances occurred during the Last glacial cycle, (2)
          what drove the different response of the individual lobes (you show
          that it is glacier response time, i.e. ice dynamics), (3) how the
          temporal dynamics of subglacial erosion was structured.}

@@

    The LGM extent of alpine glaciers has been mapped with varying level of
    across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
    Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
    Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
    into an reconstruction covering the entire Alps \citep{Ehlers.etal.2011},
    reproduced here (Fig.~4, red line) for comparison against model results.

@@

    Following the general assumption that trimlines, which mark the transition
    between the steep frost-shattered ridges and the more gentle glacially eroded
    valley troughs, represent the maximum elevation of the LGM ice surface,
    the LGM ice thickness in the Alps been reconstructed in several areas
    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
    However, this assumption is challenged by geological evidence from
    Scandinavia \citep[e.g.,]{Kleman.1994, Kleman.Borgstrom.1994,
    Stroeven.etal.2002}, the British Isles \citep[e.g.,]{Fabel.etal.2012}, and
    North America \citep[e.g.,]{Kleman.etal.2010}, that pre-glacial landscapes
    located well above the trimlines have been glaciated and preserved under
    under cold-based ice, sometimes for several glacial cycles.

@@

    Glacial history of the Alps prior to the LGM remains poorly constrained.
    Although the four glaciations model \citep{Penck.Bruckner.1909} has long
    been used, it is now known that glaciers advanced onto the foreland many
    more times \citep{Schluchter.1991, Preusser.etal.2011}. Sparse geological
    data indicate that the last glacial cycle may have comprised two or even
    three periods of glacier growth and decay \citep{Preusser.2004,Ivy-Ochs.etal.2008}.
    Luminescence dating from two sites in the central northern foreland
    indicate an early last glacial advance of Alpine glaciers onto the near
    foreland around $107\pm9$ to $101\pm5$\,ka \citep{Preusser.etal.2003,
    Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
    sparse, and its timing remain uncertain \citep[e.g.,]{Preusser.etal.2003,
    Link.Preusser.2006}.

>>>>

    For nearly 300~years, montane people and early explorers of the European
    Alps learned to read the geomorphological imprint left by glaciers in the
    landscape, and to understand that glaciers had once been more extensive
    than today \citep[e.g.,][p.~21]{Windham.Martel.1744}. Contemporaneously,
    it was also observed in the Alps that glaciers move by a combination of
    meltwater-induced \emph{sliding} at the base \citep[\S532]{Saussure.1779},
    and viscous \emph{deformation} within the ice body \citep{Forbes.1846b}. As
    glaciers flow and slide across their bed, they transport rock debris and
    erode the landscape, thereby leaving geomorphological traces of their
    former presence. In the mid-nineteenth
    century, more systematic studies of glacial features showed that alpine
    glaciers extended well below their contemporary margins \citep{Venetz.1821}
    and even onto the alpine foreland \citep{Charpentier.1841}, yielding the
    idea that, under colder temperatures, expansive ice sheets had once covered
    much of Europe and North America \citep{Agassiz.1840}.

    However, this glacial theory did not gain general acceptance until the
    discovery and exploration of the two present-day ice sheets on Earth, the
    Greenland and Antarctic ice sheets, provided a modern analogue for the
    proposed European and North American ice sheets. Although it was long
    unclear whether there had been a single or multiple \emph{glaciations},
    this controversy ended with the large-scale mapping of two distinct moraine
    systems in North America \citep{Chamberlin.1894}. In the European Alps, the
    systematic classification of the glaciofluvial terraces on the northern
    foreland indicated that there had been at least four major
    glaciations in the Alps \citep{Penck.Bruckner.1909}. More recent studies
    of glaciofluvial stratigraphy in the foreland indicate up to eight
    glaciations \citep{Ivy-Ochs.etal.2008, Preusser.etal.2011}.

    More recently, palaeoclimate records extracted from deep sea sediments and
    ice cores have provided a much more detailed picture of the Earth
    environmental history \citep[e.g.,][]{Emiliani.1955,
    Shackleton.Opdyke.1973, Dansgaard.etal.1993, Augustin.etal.2004},
    indicating neither one, nor four, but rather tens of \emph{glacial cycles}.
    During the last 800000 years (800\,ka), glacial and interglacial periods
    have succeeded each other with a 100\,ka periodicity \citep{Hays.etal.1976,
    Augustin.etal.2004}. Nevertheless, this global signal is largely governed
    by the North American and Eurasian ice sheet complexes. It is thus unclear
    wether glacier advances and retreats in the Alps were in pace with global
    sea-level fluctuations.

    Besides, the landform record is typically sparse in time and, most often,
    spatially incomplete. Palaeo-ice sheets did not leave a continuous imprint
    on the lanscape, and much of this evidence has been overprinted by
    subsequent glacier re-advances and other geomorphological processes
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2006, Kleman.etal.2010}. Dating
    uncertainties typically increase with age, such that dated reconstructions
    are strongly biased towards earlier glaciations \citep{Heyman.etal.2011}.
    The European Alps have been studied in more detail than any other glacial
    region, but they are no exception to these rules. Although sparse
    geological traces indicate that the last glacial cycle may have comprised
    two or three cycles of glacier growth and decay \citep{Preusser.2004,
    Ivy-Ochs.etal.2008}, most glacial features currently left on the foreland
    present a record of the last major glaciation of the Alps, dating from the
    \emph{Last Glacial Maximum} (LGM).

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Bini.etal.2009, Coutterand.2010,
    Husen.2011}. Although no evidence for cold-based glaciation has been found
    in the Alps to date, the assumption that trimlines, the upper limit of
    glacial erosion, represent the maximum ice surface elevation, which has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2 (29--14\,ka)
    \citep{Lisiecki.Raymo.2005}. Regional variation between different piedmont
    lobes exist \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear wether they
    relate to climate or glacier dynamics \citep{Monegato.etal.2017} or
    uncertainties in the dating methods.

    Although the glacial history of the European Alps has been studied for
    nearly three hundred years, it thus remains incompletely known
    \begin{itemize}
      \item what climate evolution lead to the known maximum ice limits,
      \item to which extent ice flow was controlled by subglacial topography,
      \item what drove the different response of the individual lobes,
      \item how far above the trimline was the ice surface located, and
      \item how many advances occurred during the last glacial cycle.
    \end{itemize}

    Here, we use the Parallel Ice Sheet Model
    \citep[PISM,][]{PISM-authors.2017}, a numerical ice sheet model that
    approximates glacier sliding and deformation (Sect.~\ref{sec:model}), to
    model alpine glacier dynamics through the last glacial cycle (120--0\,ka),
    a period for which palaeo-temperature proxies are available, albeit non
    regional. In an attempt to analyse the long-standing questions outline
    above, we test the model sensitivity to multiple palaeo-climate forcing
    (Sect.~\ref{sec:climate}), and then explore the modelled glacier dynamics
    at high resolution for the optimal forcing (Sect.~\ref{sec:results}).

@@

    The LGM extent of alpine glaciers has been mapped with varying level of
    detail across the mountain range \citep{Penck.Bruckner.1909, Jackli.1962,
    Bini.etal.2009, Coutterand.2010, Bavec.Verbic.2011,
    Buoncristiani.Campy.2011, Husen.2011}. These maps have been compiled
    into an reconstruction covering the entire Alps \citep{Ehlers.etal.2011},
    reproduced here (Fig.~4, red line) for comparison against model results.

@@

    Following the general assumption that trimlines, which mark the transition
    between the steep frost-shattered ridges and the more gentle glacially eroded
    valley troughs, represent the maximum elevation of the LGM ice surface,
    the LGM ice thickness in the Alps been reconstructed in several areas
    \citep{Husen.1987, Florineth.1998, Florineth.Schluchter.1998,
    Kelly.etal.2004, Bini.etal.2009, Coutterand.2010, Cossart.etal.2012}.
    However, this assumption is challenged by geological evidence from
    Scandinavia \citep[e.g.,][]{Kleman.1994, Kleman.Borgstrom.1994,
    Stroeven.etal.2002}, the British Isles \citep[e.g.,][]{Fabel.etal.2012}, and
    North America \citep[e.g.,][]{Kleman.etal.2010}, that pre-glacial landscapes
    located well above the trimlines have been glaciated and preserved under
    under cold-based ice, sometimes for several glacial cycles.

@@

    Glacial history of the Alps prior to the LGM remains poorly constrained.
    Although the four glaciations model \citep{Penck.Bruckner.1909} has long
    been used, it is now known that glaciers advanced onto the foreland many
    more times \citep{Schluchter.1991, Preusser.etal.2011}. Sparse geological
    data indicate that the last glacial cycle may have comprised two or even
    three periods of glacier growth and decay \citep{Preusser.2004,Ivy-Ochs.etal.2008}.
    Luminescence dating from two sites in the central northern foreland
    indicate an early last glacial advance of Alpine glaciers onto the near
    foreland around $107\pm9$ to $101\pm5$\,ka \citep{Preusser.etal.2003,
    Preusser.Schluchter.2004}. Evidence for a MIS~4 glaciation is equally
    sparse, and its timing remain uncertain \citep[e.g.,][]{Preusser.etal.2003,
    Link.Preusser.2006}.
